From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-06 paper/paper.Rmd
Revisions in progress

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  review # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \linenumbers
  \usepackage[nomarkers]{endfloat}

@@

  I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=7, fig.height=3.5)
library(tidyverse)
library(gridExtra)
library(ggthemes)
colours <- ptol_pal()(2)
```

@@

##extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())

#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
```

@@

Stochasticity is most often thought of as something which obscures
patterns we are trying to infer.  This is a simple consequence of
modeling: as discussed above, variables of interest to the processes
in question are modeled explicitly (e.g. change in population density
as it depends on the current density), everything else is just
noise.  Most statistical inference in ecology is built solely to
deal with this problem: pulling back the veil of noise to reveal
processes of interest.  Unfortunately, given the high levels of
variation and relatively limited data typical in ecological analyses,
this can be a significant challenge, and even as statistical and
computational advances provide more powerful and accessible inference
tools, the task only gets harder.

For instance, few ecological
concepts are more central than the notion of density dependence:
that competition in some form or other should limit a species ability
to grow without bound.  Yet establishing statistical evidence for
density dependence is remarkably difficult, as illustrated by
@Knape2011.  Using time series from the Global Population Dynamics
Database (GPDD) Knape & de Valpine show that once one accounts for
the potential for measurement error in addition to environmental
stochasticity -- an unavoidable source of noise in most ecological
data -- it becomes difficult to find evidence for even the simplest
models of density dependence in the majority of available time
series.  Faced with such uncertainty, it is tempting to throw up
our hands and wait until some prophesied deluge of streaming
micro-sensors stapled to every organism provide enough signal to
beat back the noise. Fortunately, this has not been the path taken.

@@

With the exception of quantum mechanics (and there only at scales
that have no relevance to ecology), stochasticity is not an inherent
property of nature but rather an emergent property of many events
occurring at a different scale.  Coins and dice obey the physical
laws of motion; only variation in how they are tossed (i.e. something
complex we do not usually model) creates randomness.  The term
"noise" itself reflects this emergent rather than fundamental origin:
the combination of hundreds of voices and sounds each individually
distinct blends together into a background of noise. Statistical
mechanics tells us that temperature is a measure of the average
speed of individual molecules in a gas -- any given molecule will
experience a range of possible speeds but by itself has no notion
of temperature.  Likewise, stochasticity enters our ecological
models because our quantities of interest, such as the density of
a population, is but an average over a lower-level, discrete process
of individual births and deaths. Stochasticity that emerges directly
from discrete individual events that underlie the process described
by our model are considered *intrinsic stochasticity*, commonly
referred to as *demographic stochasticity* in ecological literature,
as it usually arises in the context scaling from individual demographic
events to population level descriptions. An important aspect of
this demographic stochasicity is that it scales with the square
root of the "system size," so that for very large populations the
resulting fluctuations are relatively small. On the other hand,
*extrinsic stochasticity*, usually referred to as *environmental
stochasticity* in the ecological literature, arises in processes
that are not described by our model, but merely impact the value
of some parameter of our model; for instance, fluctuations in
temperature may drive variation in a death rate, or fluctuations
in rainfall drive variation in carrying capacity.

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: in a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on both the population size $N$) and increases the
population size, $N$ to $N+1$, and deaths occur with probability
$d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the "master equation."

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure 1
uses this algorithm to compare simulations of the Levins' patch model;
see Appendix A for example code.  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.

@@

Figure 1 shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

<!-- Figure 1: Gillespie simulations -->
```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system. Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A). Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2, col=colours[2]) + 
  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[2]) + 
  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[2]) + 
  geom_line(aes(y = n), col=colours[1]) +
  facet_wrap(~system_size, scales = "free_y") 
```

@@

For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
we find the average number of occupied patches 
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases only as $\sqrt{N}$ while the steady state itself increases
in direct proportion to $N$, resulting in the smaller relative size
of fluctuations in the larger system in Figure 1.  We can also see
that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with different levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

Note that the contribution of environmental stochasticity is
independent of system size and directly proportional to the variance
in the environment.  This derivation provides a useful illustration
of the difference between demographic noise and environmental noise:
the latter simply describes the variation introduced by some other
dynamic variable that we are not modelling explicitly, and whose
dynamics are independent of the state of our system, $x$. This
environmental variation no doubt also arise from some more detailed
lower level process, but we need not be explicit about those dynamics
-- the expansion tells us that it is sufficient to know the overall
variance and the degree of auto-correlation.  Recent work has also
highlighted the importance of including both demographic and
environmental noise [@Melbourne2008; @Bonsall2004].  The system
size expansion still requires the assumption that the noise is small
relative to the system size.  In the case of large fluctuations,
this approximation must be replaced with a different approach, such
as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the auto-correlation timescale
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (i.e. white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].

Other forms of uncertainty common in models are not stochastic
phenomena.  This includes any form of *reducible uncertainty*: In
constructing and estimating mathematical models of ecological
processes, we encounter uncertainty over our parameter estimates
and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for
instance, both introduce variation into any forecast of future
outcomes. Yet unlike stochasticity, this uncertainty is reducible:
with additional observations, we can narrow the uncertainty over
parameters or candidate models, while stochasticity remains an
inherent source of variation. We will this not consider model and
parameter uncertainty, as they are not sources of noise.  *Deterministic
chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the appropriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

@@

<!-- Figure 2: Quasi-cycles -->
```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.  Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations. Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
read_csv("../appendixA/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

Understanding the origin of sustained oscillations has long been a
central question in ecology [@Hastings1996; @Bjornstad2001].  The
simple Lotka Volterra model produces only neutrally stable centers,
while the more realistic model that introduces carrying capacity
for prey as they compete for resources results in a model that
produces only damped oscillations, eventually settling into a stable
state (Figure 1A).  However, the introduction of relatively small
noise into the system can be sufficient to excite the system into
sustained oscillations (Figure 1B).  In the example shown here, a
simple predator-prey system is driven by environmental white noise
$\xi_t$, with prey following the equation:

@@

$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
with $c$ the conversion ratio and $d$ the predator mortality rate.  Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$. Figure 1A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure 1B shows $\sigma_x = \sigma_y = 0.01$.

Oscillations arise through the process of stochastic resonance.
The damped oscillations seen in Figure 1a are analgous to a child
sitting still on a swing: friction slowly damps the magnitude of
each successive oscillation, while the period of pendulum remains
unchanged.  Driving the pendulum through perturbations timed to
match that natural period can quickly create sustained oscillations
with large magnitudes, but such perfect timing is not required.  In
the model above, white noise effectively drives the pendulum at all
frequencies, though at very low magnitude, since the noise is small
relative to the scale of the dynamics.  However, frequencies that
most closely match the natural period of the damped oscillator are
amplified, resonating with the natural oscillations.  This process
is sufficient to permit even small noise to drive sustained
oscillations in the model indefinitely.  The power spectrum (the
squared norm of the Fourier transform, a measure of what frequencies
are present; see, e.g. @Black2012) provides a convenient way to
visualize the effects of these oscillations.  The power spectrum
for a deterministic limit cycle would create a perfect sinusoidal
oscillation and a resulting power spectrum of a delta spike at that
single frequency.  The power spectrum for the quasi-cycle instead
shows a distribution of frequencies, centered at the the resonant
frequency of the damped oscillator.

Early work by @Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod, while @Black2010 illustrates the use of
power-spectrum approaches to examine the behavior of quasi-cycles
in common disease outbreak models, focusing on a case where noise
arises explicitly from demographic stochasticity, rather than generic
environmental noise as in the model above and elsewhere.  @Pineda-Krch2007
compare patterns from these stochastically generated cycles to
stable limit cycles they resemble.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern).

<!-- See additional citations in Coulson2004 and others that mention stochastic resonance -->

>>>>

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

@@

  I provide a brief synthesis of ecological research in stochastic phenomena.  Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5)
library(tidyverse)
library(gridExtra)
library(ggthemes)
colours <- ptol_pal()(2)
```

@@

##extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
```

@@

Stochasticity is most often thought of as something which obscures
patterns we are trying to infer.  This is a simple consequence of
modeling: variables of interest to the processes
in question are modeled explicitly (e.g. change in population density
as it depends on the current density), everything else is just
noise.  Most statistical inference in ecology is built solely to
deal with this problem: pulling back the veil of noise to reveal
processes of interest.  Unfortunately, given the high levels of
variation and relatively limited data typical in ecological analyses,
this can be a significant challenge, and even as statistical and
computational advances provide more powerful and accessible inference
tools, the task only gets harder.

For instance, few ecological
concepts are more central than the notion of density dependence:
that competition in some form or other should limit a species ability
to grow without bound.  Yet establishing statistical evidence for
density dependence is remarkably difficult, as illustrated by
@Knape2011.  Using time series from the Global Population Dynamics
Database (GPDD) Knape & de Valpine show that once one accounts for
the potential for measurement error in addition to environmental
stochasticity -- an unavoidable source of noise in most ecological
data -- it becomes difficult to find evidence for even the simplest
models of density dependence in the majority of available time
series.  Faced with such uncertainty, it is tempting to abandon
any such attempts until much longer and higher resolution data 
becomes available. Fortunately, this has not been the path taken.

@@

With the exception of quantum mechanics (and there only at scales
that have no relevance to ecology), stochasticity is not an inherent
property of nature but rather an emergent property of many events
occurring at a different scale.  Coins and dice obey the physical
laws of motion; only variation in how they are tossed (i.e. something
complex we do not usually model) creates randomness.  The term
"noise" itself reflects this emergent rather than fundamental origin:
the combination of hundreds of voices and sounds each individually
distinct blends together into a background of noise. Statistical
mechanics tells us that temperature is a measure of the average
speed of individual molecules in a gas -- any given molecule will
experience a range of possible speeds but by itself has no notion
of temperature.  Likewise, stochasticity enters our ecological
models because our quantities of interest, such as the density of
a population, is but an average over a lower-level, discrete process
of individual births and deaths. Stochasticity that emerges directly
from discrete individual events that underlie the process described
by our model are considered *intrinsic stochasticity*, commonly
referred to as *demographic stochasticity* in ecological literature,
as it usually arises in the context scaling from individual demographic
events to population level descriptions. An important aspect of
this demographic stochasicity is that it scales with the square
root of the "system size," [@vanKampen2007] so that for very large populations the
resulting fluctuations are relatively small. On the other hand,
*extrinsic stochasticity*, usually referred to as *environmental
stochasticity* in the ecological literature, arises in processes
that are not described by our model, but merely impact the value
of some parameter of our model; for instance, fluctuations in
temperature may drive variation in a death rate, or fluctuations
in rainfall drive variation in carrying capacity.

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: In a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on the population size, $N$), which increases the
population size to $N+1$. Similarly, death events occur with probability
$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the "master equation" [@vanKampen2007].

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure \ref{gillespie}
uses this algorithm to compare simulations of @Levins1969 patch model;
(see Appendix A for example code).  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.

@@

Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

<!-- Figure 1: Gillespie simulations -->
```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system. Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A). Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines. \\label{gillespie}"}
read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2, col=colours[2]) + 
  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[2]) + 
  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[2]) + 
  geom_line(aes(y = n), col=colours[1]) +
  facet_wrap(~system_size, scales = "free_y") 
```

@@

For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
we find the average number of occupied patches 
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases only as $\sqrt{N}$ while the steady state itself increases
in direct proportion to $N$, resulting in the smaller relative size
of fluctuations in the larger system in Figure \ref{gillespie}.  We can also see
that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with different levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

Note that the contribution of environmental stochasticity is
independent of system size and directly proportional to the variance
in the environment.  This derivation provides a useful illustration
of the difference between demographic noise and environmental noise:
the latter simply describes the variation introduced by some other
dynamic variable that we are not modelling explicitly, and whose
dynamics are independent of the state of our system, $x$. This
environmental variation no doubt also arises from some more detailed
lower level process, but we need not be explicit about those dynamics
-- the expansion tells us that it is sufficient to know the overall
variance and the degree of auto-correlation.  Recent work has also
highlighted the importance of including both demographic and
environmental noise [@Melbourne2008; @Bonsall2004].  The system
size expansion still requires the assumption that the noise is small
relative to the system size.  In the case of large fluctuations,
this approximation must be replaced with a different approach, such
as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the auto-correlation timescale
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (i.e. white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].

Other forms of uncertainty common in models are not stochastic
phenomena.  This includes any form of *reducible uncertainty*: In
constructing and estimating mathematical models of ecological
processes, we encounter uncertainty over our parameter estimates
and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for
instance, both introduce variation into any forecast of future
outcomes. Yet unlike stochasticity, this uncertainty is reducible:
with additional observations, we can narrow the uncertainty over
parameters or candidate models. Thus we will not consider model and
parameter uncertainty, as they do not reflect intrinsic phenomena of 
the dynamics but only our own ignorance about those dynamics.  
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the appropriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

@@

<!-- Figure 2: Quasi-cycles -->
```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.  Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations. Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A. \\label{quasicycles}"}
read_csv("../appendixA/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

Understanding the origin of sustained oscillations has long been a
central question in ecology [@Hastings1996; @Bjornstad2001].  The
simple Lotka Volterra model produces only neutrally stable centers,
while the more realistic model that introduces carrying capacity
for prey as they compete for resources results in a model that
produces only damped oscillations, eventually settling into a stable
state (Figure \ref{quasicycles}A).  However, the introduction of relatively small
noise into the system can be sufficient to excite the system into
sustained oscillations (Figure \ref{quasicycles}B).  In the example shown here, a
simple predator-prey system is driven by environmental white noise
$\xi_t$, with prey following the equation:

@@

$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
with $c$ the conversion ratio and $d$ the predator mortality rate.  Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$. Figure \ref{quasicycles}A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure \ref{quasicycles}B shows $\sigma_x = \sigma_y = 0.01$.

Oscillations arise through the process of stochastic resonance.
The damped oscillations seen in Figure \ref{quasicycles}A are analgous to a child
sitting still on a swing: friction slowly damps the magnitude of
each successive oscillation, while the period of pendulum remains
unchanged.  Driving the pendulum through perturbations timed to
match that natural period can quickly create sustained oscillations
with large magnitudes, but such perfect timing is not required.  In
the model above, white noise effectively drives the pendulum at all
frequencies, though at very low magnitude, since the noise is small
relative to the scale of the dynamics.  However, frequencies that
most closely match the natural period of the damped oscillator are
amplified, resonating with the natural oscillations.  This process
is sufficient to permit even small noise to drive sustained
oscillations in the model indefinitely.  The power spectrum (the
squared norm of the Fourier transform, a measure of what frequencies
are present; see, e.g. @Black2012) provides a convenient way to
visualize the effects of these oscillations.  The power spectrum
for a deterministic limit cycle would create a perfect sinusoidal
oscillation and a resulting power spectrum of a delta spike at that
single frequency.  The power spectrum for the quasi-cycle instead
shows a distribution of frequencies, centered at the the resonant
frequency of the damped oscillator.

Early work by @Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod, while @Black2010 illustrates the use of
power-spectrum approaches to examine the behavior of quasi-cycles
in common disease outbreak models, focusing on a case where noise
arises explicitly from demographic stochasticity, rather than generic
environmental noise as in the model above and elsewhere.  @Pineda-Krch2007
compare patterns from these stochastically generated cycles to
stable limit cycles they resemble.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

<!-- See additional citations in Coulson2004 and others that mention stochastic resonance --> 
<!-- [@Keeling2001; @Coulson2004;]  -->
